Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-13 paper.Rmd
Reduce the Model Specification section

Deep learning models are in general very sensitive to the input data. The quality and relevance of the input data greatly influence the model's capacity to generate insightful and meaningful results. There are several designs of the input format that can be considered.

A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the variable input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One strategy is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length. Alternatively, sampling a fixed number of residual-fitted value pairs can ensure consistent input size. By carefully designing the sampling process, it is possible to maintain the input's representativeness while conforming to a fixed format.

Another architecture of the input layer involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. Encoding data points as pixels introduces certain trade-offs. While some detail may be lost during the discretization process, this approach aligns with the sampling-based strategy by focusing on summarizing the overall data distribution. Unlike vectors, which struggle to represent spatial relationships, image-based inputs naturally encode the joint distribution of residuals and fitted values in a visually interpretable format. This richer representation gives the model a more comprehensive and intuitive understanding of the residual plot's underlying structure. Key considerations for using image-based input include image resolution and the aesthetics of the residual plot. In general, a higher resolution provides more information to the model but comes at the cost of of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option. @chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar. Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering true residual plots to be distinct from null plots of any fitted regression model. A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$. It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples. For this setup, we can apply the same criteria to define similarity between images. However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol. However, as the number of residual plots in a lineup increases, the resolution of the input image grows rapidly, posing challenges in training the model. We experimented with this approach in a pilot study, but the performance of the trained model was sub-optimal.

Taking into account the implementation cost and the need for model interpretability, we used the single residual plot input format in this paper.

@@

Given that the input is a single residual plot represented as a fixed-resolution image, the output of the computer vision model can be designed for various tasks: binary classification, categorical classification with multiple classes, or numeric regression.

The binary outcome can represent whether the input image is consistent with a null plot as determined by either (1) the data generating process or (2) the result of a visual test based on human judgement. Training a model following the latter option requires data from prior human subject experiments, presenting difficulties in controlling the quality of data due to variations in experimental settings across different studies. Additionally, some visual inference experiments are unrelated to linear regression models or residual plot diagnostics, resulting in a limited amount of available training data.

For a multi-class outcome, the model could classify images into categories such as "contains outliers" vs. "does not contain outliers" or "is non-linear" vs. "not non-linear," among other diagnostic features. While this approach is appealing in theory, it essentially reverts to the traditional testing framework, where tests focus on detecting specific types of model violations in isolation.

One significant advantage of the lineup protocol for residual diagnostics is its ability to detect multiple model violations simultaneously, provided these violations produce visually distinguishable differences in the plots. In contrast, training a classification model would require explicitly defining and mapping all potential violations of interest, or at least the most common ones, which can be impractical. If the goal is to identify specific cases, conventional statistical tests grounded in regression model assumptions are often more effective. Computer vision models may be less suited to such tasks.

Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues. However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined. For the purpose of training a model, this numeric measure has to be quantifiable.

In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot. This is further explained in \ref{sec-model-distance-estimation}. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

>>>>

Deep learning models are sensitive to input design, and several alternatives are available for encoding residual plots.

A simple approach involves feeding vectors of residuals and fitted values into the model. While this format contains all relevant information, the input length varies with sample size, which conflicts with the fixed-shape requirements of modern vision models such as those implemented in TensorFlow [@abadi2016tensorflow]. Padding can be used to enforce uniform length but risks truncation if inputs exceed the fixed size. Alternatively, fixed-size sampling of residual–fitted value pairs can ensure consistency, though at the cost of some information loss.

Another strategy is to convert residual plots into images. This enables the use of standard convolutional architectures such as VGG16 [@simonyan2014very], which can effectively capture spatial patterns. Although discretization introduces some losses in detail, image-based inputs offer a richer representation by preserving the joint distribution of residuals and fitted values in a visually interpretable format. To ensure model generalization, it is essential to use residual plots in a consistent style, i.e., same aesthetics, layout, scaling and color schemes.

Using multiple residual plots as input, such as pairs, triplets, or full lineups is also a possible option. For instance, triplet networks [@chopra2005learning] can assess visual similarity if we assume null plots are more similar from each other than from the true residual plot, which may help distinguish null plots from true residual plots. However, these architectures introduce complexity in training due to weight sharing and specialized loss functions. We experimented with this setting in a pilot study, and found multiple residual plots led to high input resolution, high computational cost and suboptimal model performance.

Balancing accuracy, interpretability, and implementation cost, we adopt a single residual plot in image format as the model input.

@@

Given the input is a single fixed-resolution image, the model output can be defined for various tasks, including binary classification, multiclass classification, and numeric regression.

In the binary case, the outcome may indicate whether the input image is consistent with a null plot, as determined either by (1) the data-generating process or (2) the outcome of a visual test based on human judgment. The first approach models the probability that a residual plot is not a null plot and is a viable option. In contrast, the second relies on experimental data that are often scarce or inconsistent across studies.

Alternatively, a multiclass outcome may classify images into categories such as "contains outliers" vs. "does not contain outliers" or "is non-linear" vs. "not non-linear," among other diagnostic features. While this approach is appealing in theory, it essentially reverts to the traditional testing framework, where tests focus on detecting specific types of model violations in isolation.

A third option is to produce a meaninful and interpretable numerical measure that quantifies the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues. However, such measures are often used informally in daily communications but rarely formalized. For supervised learning, they must be clearly defined and reliably estimable.

In this study, we defined and used a continuous distance between a true residual plot and a theoretically "good" residual plot. This approach captures the degree to which a plot deviates from model assumptions. In contrast, a probability output based on binary labels, where true residual plots are labeled as 1 and null plots as 0, cannot express the severity of violation. It forces the model to infer this information solely from the data distribution, which may be inefficient and discards useful prior knowledge.

### 2025-06-13 paper.tex
Reduce the Model Specification section

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results. There
are several designs of the input format that can be considered.

A straightforward architecture of the input layer involves feeding a
vector of residuals along with a vector of fitted values, essentially
providing all the necessary information for creating a residuals vs
fitted values plot. However, a drawback of this method is the variable
input size, which changes based on the number of observations. For
modern computer vision models implemented in mainstream software like
TensorFlow \citep{abadi2016tensorflow}, the input shape is typically
fixed. One strategy is to pad the input vectors with leading or trailing
zeros when the input tensor expects longer vectors, but it may fail if
the input vector surpasses the designed length. Alternatively, sampling
a fixed number of residual-fitted value pairs can ensure consistent
input size. By carefully designing the sampling process, it is possible
to maintain the input's representativeness while conforming to a fixed
format.

Another architecture of the input layer involves using an image as
input. The primary advantage of this design, as opposed to the vector
format, is the availability of the existing and sophisticated image
processing architectures developed over the years, such as the VGG16
architecture proposed in \citet{simonyan2014very}. These architectures
can effectively capture and summarize spatial information from nearby
pixels, which is less straightforward with vector input. Encoding data
points as pixels introduces certain trade-offs. While some detail may be
lost during the discretization process, this approach aligns with the
sampling-based strategy by focusing on summarizing the overall data
distribution. Unlike vectors, which struggle to represent spatial
relationships, image-based inputs naturally encode the joint
distribution of residuals and fitted values in a visually interpretable
format. This richer representation gives the model a more comprehensive
and intuitive understanding of the residual plot's underlying structure.
Key considerations for using image-based input include image resolution
and the aesthetics of the residual plot. In general, a higher resolution
provides more information to the model but comes at the cost of of
increased complexity and greater difficulty in training. As for the
aesthetics of the residual plot, a practical solution is to consistently
present residual plots in the same style to the model. This implies that
the model can not accept arbitrary images as input but requires the use
of the same pre-processing pipeline to convert residuals and fitted
values into a standardized-style residual plot.

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering true residual plots to be distinct from null plots of
any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

Presenting a lineup to a model aligns closely with the lineup protocol.
However, as the number of residual plots in a lineup increases, the
resolution of the input image grows rapidly, posing challenges in
training the model. We experimented with this approach in a pilot study,
but the performance of the trained model was sub-optimal.

Taking into account the implementation cost and the need for model
interpretability, we used the single residual plot input format in this
paper.

@@

Given that the input is a single residual plot represented as a
fixed-resolution image, the output of the computer vision model can be
designed for various tasks: binary classification, categorical
classification with multiple classes, or numeric regression.

The binary outcome can represent whether the input image is consistent
with a null plot as determined by either (1) the data generating process
or (2) the result of a visual test based on human judgement. Training a
model following the latter option requires data from prior human subject
experiments, presenting difficulties in controlling the quality of data
due to variations in experimental settings across different studies.
Additionally, some visual inference experiments are unrelated to linear
regression models or residual plot diagnostics, resulting in a limited
amount of available training data.

For a multi-class outcome, the model could classify images into
categories such as ``contains outliers'' vs.~``does not contain
outliers'' or ``is non-linear'' vs.~``not non-linear,'' among other
diagnostic features. While this approach is appealing in theory, it
essentially reverts to the traditional testing framework, where tests
focus on detecting specific types of model violations in isolation.

One significant advantage of the lineup protocol for residual
diagnostics is its ability to detect multiple model violations
simultaneously, provided these violations produce visually
distinguishable differences in the plots. In contrast, training a
classification model would require explicitly defining and mapping all
potential violations of interest, or at least the most common ones,
which can be impractical. If the goal is to identify specific cases,
conventional statistical tests grounded in regression model assumptions
are often more effective. Computer vision models may be less suited to
such tasks.

Alternatively, the output could be a meaningful and interpretable
numerical measure useful for assessing residual plots, such as the
strength of suspicious visual patterns reflecting the extent of model
violations, or the difficulty index for identifying whether a residual
plot has no issues. However, these numeric measures are often informally
used in daily communication but are not typically formalized or
rigorously defined. For the purpose of training a model, this numeric
measure has to be quantifiable.

In this study, we chose to define and use a distance between a true
residual plot and a theoretically ``good'' residual plot. This is
further explained in \ref{sec-model-distance-estimation}.
\citet{vo2016localizing} have also demonstrated that defining a proper
distance between images can enhance the matching accuracy in image
search compared to a binary outcome model.

>>>>

Deep learning models are sensitive to input design, and several
alternatives are available for encoding residual plots.

A simple approach involves feeding vectors of residuals and fitted
values into the model. While this format contains all relevant
information, the input length varies with sample size, which conflicts
with the fixed-shape requirements of modern vision models such as those
implemented in TensorFlow \citep{abadi2016tensorflow}. Padding can be
used to enforce uniform length but risks truncation if inputs exceed the
fixed size. Alternatively, fixed-size sampling of residual--fitted value
pairs can ensure consistency, though at the cost of some information
loss.

Another strategy is to convert residual plots into images. This enables
the use of standard convolutional architectures such as VGG16
\citep{simonyan2014very}, which can effectively capture spatial
patterns. Although discretization introduces some losses in detail,
image-based inputs offer a richer representation by preserving the joint
distribution of residuals and fitted values in a visually interpretable
format. To ensure model generalization, it is essential to use residual
plots in a consistent style, i.e., same aesthetics, layout, scaling and
color schemes.

Using multiple residual plots as input, such as pairs, triplets, or full
lineups is also a possible option. For instance, triplet networks
\citep{chopra2005learning} can assess visual similarity if we assume
null plots are more similar from each other than from the true residual
plot, which may help distinguish null plots from true residual plots.
However, these architectures introduce complexity in training due to
weight sharing and specialized loss functions. We experimented with this
setting in a pilot study, and found multiple residual plots led to high
input resolution, high computational cost and suboptimal model
performance.

Balancing accuracy, interpretability, and implementation cost, we adopt
a single residual plot in image format as the model input.

@@

Given the input is a single fixed-resolution image, the model output can
be defined for various tasks, including binary classification,
multiclass classification, or numeric regression.

In the binary case, the outcome may indicate whether the input image is
consistent with a null plot, as determined either by (1) the
data-generating process or (2) the outcome of a visual test based on
human judgment. The first approach models the probability that a
residual plot is not a null plot and is a viable option. In contrast,
the second relies on experimental data that are often scarce or
inconsistent across studies.

Alternatively, a multiclass outcome may classify images into categories
such as ``contains outliers'' vs.~``does not contain outliers'' or ``is
non-linear'' vs.~``not non-linear,'' among other diagnostic features.
While this approach is appealing in theory, it essentially reverts to
the traditional testing framework, where tests focus on detecting
specific types of model violations in isolation.

A third option is to produce a meaninful and interpretable numerical
measure that quantifies the strength of suspicious visual patterns
reflecting the extent of model violations, or the difficulty index for
identifying whether a residual plot has no issues. However, such
measures are often used informally in daily communications but rarely
formalized. For supervised learning, they must be clearly defined and
reliably estimable.

In this study, we defined and used a continuous distance between a true
residual plot and a theoretically ``good'' residual plot. This approach
captures the degree to which a plot deviates from model assumptions. In
contrast, a probability output based on binary labels, where true
residual plots are labeled as 1 and null plots as 0, cannot express the
severity of violation. It forces the model to infer this information
solely from the data distribution, which may be inefficient and discards
useful prior knowledge.
